\Large{Adversarial Training against Systematic Uncertainty}

### 2016-09-12 tex/paper.tex
save

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary model $r := p_{\theta_r}(z |
f(X;\theta_f))$ of parameters $\theta_r$ and associated loss ${\cal
L}_r(\theta_f, \theta_r)$. This model takes  as input realizations of $f(X;
\theta_f)$, for the current value $\theta_f$ of $f$ parameters, and produces as
output a function $p_{\theta_r}(z | f(X;\theta_f))$ modeling the conditional probability density that $f(X;
\theta_f)$ is generated from the nuisance value $z$.
\glnote{Give examples of $r$ for categorical and continuous nuisances.}
Intuitively, if $p(f(X;
\theta_f)|z)$ varies with $z$, then the corresponding correlation can be
captured by $r$. By contrast, if $p(f(X; \theta_f)|z)$ is invariant with $z$, as
we require, then $r$ should perform poorly and be close to random guessing.
Training $f$ such that it additionally minimizes the performance of $r$
therefore acts as a regularization towards Eqn.~\ref{eqn:criterion}.

@@

\glnote{Add summary figure.}

@@

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty
\glnote{add ref} by estimating $p(f(x)|z)$ with a parameterized calibration
procedure. Clearly, this classifier is however not optimal for $z \neq z_0$. In
this setting, parameterized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, such classifiers can however not be used on real data since
the correct value $z$ of the nuisance often remains unknown. This is typically
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
marginalized out, but otherwise often limits the range of their applications. In
practice, parameterized classifiers  are also often computationally expensive to
build and evaluate. In particular, calibrating their decision function, i.e.
approximating $p(f(x,z)|z)$ as a continuous function of $z$, remains an open
challenge. By contrast, constraining $f$ to be pivotal yields a classifier which
may not be optimal with respect to $Y|X,Z$, as discussed in Section~\ref{sec:theory}, but that can be
used in a wider range of applications, since knowing the correct value $z$ of
the nuisance is not necessary, and for which calibration needs to be carried out
only once, since  the dependence on the nuisance is now built in the classifier.
\glnote{Shall we also discuss the relation with \citep{Neal:2007zz} where point
estimates of the nuisance are used as inputs to $f$?}

In machine learning, learning a pivotal quantity can be related to the problem of domain
adaptation
\citep{blitzer2006domain,pan2011domain,gopalan2011domain,gong2013connecting,baktashmotlagh2013unsupervised,ganin2014unsupervised},
where the goal is often stated as trying to learn a domain-invariant
representation of the data. Similarly, our method also relates to the problem of
enforcing fairness in classification \citep{zemel2013learning,EdwardsS15}. For
both families of methods, the problem can equivalently be stated as learning a
classifier which is a pivotal quantity with respect to either the domain or some
chosen attribute of the data, such as gender, color or age. In this context,
\citep{ganin2014unsupervised,EdwardsS15} are certainly among the closest to our
work, in which domain invariance and fairness are enforced through an adversarial
minimax setup composed of a classifier and an adversary discriminator. In
particular, our work can be regarded as a generalization that also supports the
continuous case, which can be viewed as handling infinitely many domains,
provided they can be continuously parameterized, or as enforcing fairness over
continuous attributes.

>>>>

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary model $r := p_{\theta_r}(z |
f(X;\theta_f))$ of parameters $\theta_r$ and associated loss ${\cal
L}_r(\theta_f, \theta_r)$. This model takes  as input realizations of $f(X;
\theta_f)$, for the current value $\theta_f$ of $f$ parameters, and produces as
output a function $p_{\theta_r}(z | f(X;\theta_f))$ modeling the posterior
probability density that $z$ parameterizes the sample $X$ observed through
$f(\cdot;\theta_f)$. If $Z$ takes discrete values, then $p_{\theta_r}$ can be
represented as a probabilistic classifier $\mathbb{R} \mapsto \mathbb{R}^{|{\cal
Z|}}$ whose output $j$ (for $j=1, \dots, |{\cal Z}|$) is the estimated
probability mass $p_{\theta_r}(z_j|f(X;\theta_f))$. Similarly, if $Z$ takes
continuous values and if we assume some parameteric distribution for
$Z|f(X;\theta_f)$ (e.g., a gaussian), then $p_{\theta_r}$ can be represented as
network whose output $j$ is the estimated value of the corresponding parameter
of that distribution (e.g., its mean and variance), and such that
$p_{\theta_r}(z|f(X;\theta_f))$ can then be evaluated for any $z \in {\cal Z}$
under that distribution.  Intuitively, if $p(f(X; \theta_f)|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)|z)$ is invariant with $z$, as we require, then $r$ should
perform poorly and be close to random guessing. Training $f$ such that it
additionally minimizes the performance of $r$ therefore acts as a regularization
towards Eqn.~\ref{eqn:criterion}.

@@

\glnote{Discuss that we could equivalently require an intermediate representation of the data (as learned internally by $f$) to be pivotal, instead of requiring $f(X)$ to be pivotal.}
\glnote{Add summary figure.}

@@

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty
\glnote{add ref} by estimating $p(f(x)|z)$ with a parameterized calibration
procedure. Clearly, this classifier is however not optimal for $z \neq z_0$. In
this setting, parameterized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, such classifiers can however not be used on real data since
the correct value $z$ of the nuisance often remains unknown. This is typically
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
marginalized out, but otherwise often limits the range of their applications. In
practice, parameterized classifiers  are also computationally expensive to build
and evaluate. In particular, calibrating their decision function, i.e.
approximating $p(f(x,z)|z)$ as a continuous function of $z$, remains an open
challenge. By contrast, constraining $f$ to be pivotal yields a classifier which
may not be optimal with respect to $Y|X,Z$, as discussed in
Section~\ref{sec:theory}, but that can otherwise be used in a wider range of
applications, since knowing the correct value $z$ of the nuisance is not
necessary. Similarly, calibration needs to be carried out only once, since  the
dependence on the nuisance is now built-in. \glnote{Shall we also discuss the
relation with \citep{Neal:2007zz} where point estimates of the nuisance are used
as inputs to $f$?}

In machine learning, learning a pivotal quantity can be related to the problem
of domain adaptation
\citep{blitzer2006domain,pan2011domain,gopalan2011domain,gong2013connecting,baktashmotlagh2013unsupervised,ganin2014unsupervised},
where the goal is often stated as trying to learn a domain-invariant
representation of the data. Likewise, our method also relates to the problem of
enforcing fairness in classification \citep{zemel2013learning,EdwardsS15}, which
is stated as learning a classifier that is independent of some chosen attribute
such as gender, color or age. For both families of methods, the problem can
equivalently be stated as learning a classifier which is a pivotal quantity with
respect to either the domain or the selected feature. In this context,
\citep{ganin2014unsupervised,EdwardsS15} are certainly among the closest to our
work, in which domain invariance and fairness are enforced through an
adversarial minimax setup composed of a classifier and an adversary
discriminator. Following this line of work, our method can be regarded as a
generalization that also supports the continuous case, which can be viewed as
handling infinitely many domains, provided they can be continuously
parameterized, or as enforcing fairness over continuous attributes.
